Adaptive Waveform Inversion for Transmitted Wave Data

### 2025-10-28 repro/papers/awi/paper.tex
minor

The version of AWI introduced by \cite{Warner:16} is based on an
active source acoustic model of data generation in Euclidean space
$\bR^3$.

@@

The parameter to be determined in the inverse problem studied here is
the bulk modulus $\kappa$. It is presumed is bounded and smooth:
\begin{equation}
  \label{eqn:coeff}
  \log \kappa \in C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
\end{equation}
Under these hypotheses, the symmetric hyperbolic system \ref{eqn:awe}
has a unique distribution solution (for example \cite{Lax:PDENotes},
Theorem 7.4). The Propagation of
Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) %and
%the pull-back property of distributions (\cite{Dui:95}, Proposition
%1.3.3)
imples that $p(\cdot,\cdot;\bx_s)$ is smooth away from $\{\bx_s\}
\times \bR$, so that the traces $p(\bx_r, \cdot;\bx_s)$ on
$\{(\bx_r,t): t \in \bR\}$ are well-defined.

@@

The analysis to follow requires three further constraints on the set
of source-receiver location pairs $X_{sr}$ and the domain
$M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ of $F$. The first is
the {\em Exponential Decay} assumption:
\begin{quote}
  For any $n > 0$, $f \in C_0^{\infty}(\bR)$,
  there exist $K >
  0, T \in \bR,$ and $\delta>0$ so
  that for any $k$ with $0 \le k \le n$, $(\bx_s,\bx_r) \in
 X_{sr},$ and any $\log \kappa \in M$, the solution $(p,\bv)$ of \ref{eqn:awe} is smooth
  for $t>T$ and satisfies
  \[
    |\partial_t^k p(\bx_r,t;\bx_s)| \le K e^{-\delta t}, \, t > T.
  \]
\end{quote}
SInce $X_{sr}$ is bounded, the property described in this assumption
is a special case of {\em local energy decay}, a much studied topic. The {\em
  non-trapping} hypothesis is sufficient to ensure local energy decay:
all rays of geometric asymptotics (that is, geodesics for the Riemannian metric associated
to the system \ref{eqn:awe}) escape to infinity (\cite{Hristova:09},
Theorem 2, \cite{EgorovShubin}, Theorem 2.104, p. 192). That is,
exponential decay follows from a characteristic of $\kappa$.

@@

  \begin{itemize}
  \item[1. ] for $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, and
    $\sigma, \lambda >0$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[2. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[3. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

@@

\subsection{Hadamard's Decomposition}
The Hadamard decomposition of the acoustic Green's function \cite[]{Friedlander:75,Qian:JCP24} provides a
basis for analysis of AWI and MSWI. As presented by
\cite{Friedlander:75}, for instance, this construction applies to
second-order wave equations, but the system \ref{eqn:awe} is
equivalent fo a second-order problem: differentiate the first equation
with respect to $t$, eliminate $\bv$ via the second equation, and
rearrange to obtain
\begin{equation}
  \label{eqn:awe2}
  \frac{\rho}{\kappa}\frac{\partial^2 p}{\partial t^2} - \nabla^2 p -
  \nabla \log \rho \cdot \nabla p = \rho \frac{\partial f}{\partial t}
  \delta(\bx-\bx_s), \nonumber \\
  p=0, \, t \ll 0.
\end{equation}
If $(p,\bv)$ is a distribution solutions of the system
\ref{eqn:awe} then $p$ is a solution of equation
\ref{eqn:awe2}, as follows by differentiating the first equation of
\ref{eqn:awe} with respect to $t$ and eliminating $\bv$ via the second
equation. Conversely, if $p$ is a solution of the second-order initial
value problem\ref{eqn:awe2}, then under the conditions imposed in the
previous section, those to be imposed below),
\[
  \bv = \beta \int_{-\infty}^t p
\]
defines a distribution, and $(p,\bv)$ is a weak solutino of the system \ref{eqn:awe}.

@@

\begin{prop}
  \label{thm:leading}
In the definitions \ref{eqn:mswidefmain} of $J_{\rm MSWI,\sigma}$ and
\ref{eqn:awidefmain} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
(equation \ref{eqn:lamsig})
and replace the data $d$ with
$\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$, and $\ul$
with the solution $\uzl$ of the corresponding least squares problem. That is,
\begin{eqnarray}
  \label{eqn:leading}
  \dla & = & \Fla[\kappa^*] = a[\kappa^*]\delta_{\tau[\kappa^*]}*\wl
             \nonumber \\
  \Sla[\kappa]u &=& a[\kappa]\delta_{\tau[\kappa^*]}*\wl *u\nonumber
  \\
  \uzl &=& \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|\Sla[\kappa]u-\dla\|^2 + r\lambda\|u\|^2)
\end{eqnarray}
%and
%\begin{eqnarray}
%  \label{eqn:awileading}
%  J^0_{\rm MSWI,\lambda}[\kappa;\dla] & = &
%                                         \|T\uzl[\kappa;\dla]\|^2\nonumber\\
%  J^0_{\rm AWI,\lambda}[\kappa;\dla] & = &
%                                       \sum_{(\bx_s,\bx_r) \in X_{sr}} \frac{\|T\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}{\|\uzl[%\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}
%\end{eqnarray}                 
Then there exists $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
\rightarrow \bR^+$ so that for
$\kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$, $C_* \le
W[\kappa](\bx_s,\bx_r) \le C^*$, and $\uzl$ defined in display
\ref{eqn:leading} for $0 < \lambda \le 1$, 
\begin{equation}
  \label{eqn:mswittz}
 | \|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2- 
  \frac{1}{\lambda} W[\kappa](\bx_s,\bx_r) (\tau[\kappa^*]({\bf
  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 | \le C\lambda,
\end{equation}
and
\begin{equation}
  \label{eqn:awittz}
  \left| \frac{\|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2}{
        \|\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2} - (\tau[\kappa^*](\bx_s,\bx_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2\right|  \le C\lambda^2.
\end{equation}
Finally, for $\sigma = r\lambda$ with $r$ sufficiently small,
\begin{equation}
  \label{eqn:Sutz}
  \Sla[\kappa]\uzl[\kappa;\dzl] \ge \frac{1}{2}\|\dzl\|
\end{equation}
\end{prop}

>>>>

As introduced by \cite{Warner:16}, AWI is based on an
active source acoustic model of data generation in Euclidean space
$\bR^3$.

@@

The parameter to be determined in the inverse problem studied here is
the bulk modulus $\kappa$. It is presumed is bounded and smooth: $\log
\kappa \in M$, where $M$ is a bounded subset of $C^{\infty}(\bR^3) \cap
L^{\infty}(\bR^3)$, on which further constraints will be placed by and
by.

Under these hypotheses, the symmetric hyperbolic system \ref{eqn:awe}
has a unique distribution solution (for example \cite{Lax:PDENotes},
Theorem 7.4). The Propagation of
Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) %and
%the pull-back property of distributions (\cite{Dui:95}, Proposition
%1.3.3)
imples that $p(\cdot,\cdot;\bx_s)$ is smooth away from $\{\bx_s\}
\times \bR$, so that the traces $p(\bx_r, \cdot;\bx_s)$ on
$\{(\bx_r,t): t \in \bR\}$ are well-defined.

@@

The analysis to follow requires three further constraints on the set
of source-receiver location pairs $X_{sr}$ and the domain
$M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ of $F$. The first is
the {\em Exponential Decay} assumption:
\begin{quote}
  For any $n > 0$, $f \in C_0^{\infty}(\bR)$,
  there exist $K >
  0, T \in \bR,$ and $\delta>0$ so
  that for any $k$ with $0 \le k \le n$, $(\bx_s,\bx_r) \in
 X_{sr},$ and any $\log \kappa \in M$, the solution $(p,\bv)$ of \ref{eqn:awe} is smooth
  for $t>T$ and satisfies
  \[
    |\partial_t^k p(\bx_r,t;\bx_s)| \le K e^{-\delta t}, \, t > T.
  \]
\end{quote}
SInce $X_{sr}$ is bounded, the property described in this assumption
is a special case of {\em local energy decay}, a much studied topic. The {\em
  non-trapping} hypothesis is sufficient to ensure local energy decay:
all rays of geometric asymptotics (that is, geodesics for the Riemannian metric associated
to the system \ref{eqn:awe}) escape to infinity (\cite{Hristova:09},
Theorem 2, \cite{EgorovShubin}, Theorem 2.104, p. 192). In particular,
the Exponential Decay assumption is satisfied if for any $R > r>0$ and $\log
\kappa \in M$, there exists $T>0$ so that any ray ${\bf X}:\bR^+\rightarrow \bR^3$
  with ${\bf X}(0) \in B_r(0)$ satisfies $|{\bf X}(t)|>R$ for $t > T$. That is,
exponential decay may follow from a propery of $M$.

@@

  \begin{itemize}
  \item[1. ] for $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, and
    $\sigma, \lambda >0$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[3. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[4. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

@@

\subsection{Hadamard's Decomposition}
The Hadamard decomposition of the acoustic Green's function \cite[]{Friedlander:75,Qian:JCP24} provides a
basis for analysis of AWI and MSWI. As presented by
\cite{Friedlander:75}, for instance, this construction applies to
second-order wave equations, but the system \ref{eqn:awe} is
equivalent fo a second-order problem: differentiate the first equation
with respect to $t$, eliminate $\bv$ via the second equation, and
rearrange to obtain
\begin{equation}
  \label{eqn:awe2}
  \frac{\rho}{\kappa}\frac{\partial^2 p}{\partial t^2} - \nabla^2 p -
  \nabla \log \rho \cdot \nabla p = \rho \frac{\partial f}{\partial t}
  \delta(\bx-\bx_s), \nonumber \\
  p=0, \, t \ll 0.
\end{equation}
If $(p,\bv)$ is a distribution solutions of the system
\ref{eqn:awe} then $p$ is a solution of equation
\ref{eqn:awe2}, as follows by differentiating the first equation of
\ref{eqn:awe} with respect to $t$ and eliminating $\bv$ via the second
equation. Conversely, if $p$ is a solution of the second-order initial
value problem\ref{eqn:awe2}, then under the conditions imposed in the
previous section,
\[
  \bv = \beta \int_{-\infty}^t \nabla p
\]
defines a distribution, and $(p,\bv)$ is a weak solutino of the system \ref{eqn:awe}.

@@

\begin{prop}
  \label{thm:leading}
In the definitions \ref{eqn:mswidefmain} of $J_{\rm MSWI,\sigma}$ and
\ref{eqn:awidefmain} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
(equation \ref{eqn:lamsig})
and replace the data $d$ with
$\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$, and $\ul$
with the solution $\uzl$ of the corresponding least squares problem. That is,
\begin{eqnarray}
  \label{eqn:leading}
  \dla & = & \Fla[\kappa^*] = a[\kappa^*]\delta_{\tau[\kappa^*]}*\wl
             \nonumber \\
  \Sla[\kappa]u &=& a[\kappa]\delta_{\tau[\kappa^*]}*\wl *u\nonumber
  \\
  \uzl &=& \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|\Sla[\kappa]u-\dla\|^2 + r\lambda\|u\|^2)
\end{eqnarray}
%and
%\begin{eqnarray}
%  \label{eqn:awileading}
%  J^0_{\rm MSWI,\lambda}[\kappa;\dla] & = &
%                                         \|T\uzl[\kappa;\dla]\|^2\nonumber\\
%  J^0_{\rm AWI,\lambda}[\kappa;\dla] & = &
%                                       \sum_{(\bx_s,\bx_r) \in X_{sr}} \frac{\|T\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}{\|\uzl[%\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}
%\end{eqnarray}                 
Then there exists $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
\rightarrow \bR^+$ so that for
$\kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$, $C_* \le
W[\kappa](\bx_s,\bx_r) \le C^*$, and $\uzl$ defined in display
\ref{eqn:leading} for $0 < \lambda \le 1$, 
\begin{equation}
  \label{eqn:mswittz}
 | \|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2- 
  \frac{1}{\lambda} W[\kappa](\bx_s,\bx_r) (\tau[\kappa^*]({\bf
  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 | \le C\lambda,
\end{equation}
and
\begin{equation}
  \label{eqn:awittz}
  \left| \frac{\|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2}{
        \|\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2} - (\tau[\kappa^*](\bx_s,\bx_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2\right|  \le C\lambda^2.
\end{equation}
Finally, for $\sigma = r\lambda$ with $r$ sufficiently small,
\begin{equation}
  \label{eqn:Sutz}
  \|\Sla[\kappa]\uzl[\kappa;\dzl]\| \ge \frac{1}{2}\|\dzl\|
\end{equation}
\end{prop}
